---
title: "Transport variable explore"
author: "Mike Spencer"
date: "09/12/2021"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = F, warning = F,
                      fig.width = 10, fig.height = 6)
options(scipen = 999)
```

```{r packages}
library(tidyverse)
library(haven)
library(knitr)
```

```{r plot theme}
theme_temp = function(){
  theme_bw() +
    theme(text = element_text(size = 15))
}
```


## Intro

This is an RMarkdown document, summarising variables of interest.
I'm considering the relationship between income and mode of travel.
For example, is cycling the preserve of the affluent middle class?

Understanding Society variable guide: <https://www.understandingsociety.ac.uk/documentation/mainstage/dataset-documentation?search_api_views_fulltext=salary>.


## Data prep

```{r data}
df = read_dta("~/Cloud/personal/gofcoe/understanding_society/6614stata_B17CC6790677EF32F72CE50881AE98E1B9FC1F79133B07B63B353396D3AB917A_V1/UKDA-6614-stata/stata/stata13_se/ukhls_w10/j_indresp.dta") %>% 
  select(pidp, j_pdvage,
         contains("wktrv"), j_workdis,
         j_fimnnet_dv,
         j_benbase1, j_benbase2, j_benbase4)
```

```{r data transport, eval = F}
tran_opt = tibble(name = paste0("j_wktrv", c(1:10, 97)),
       val = c("Drive myself", "Get a lift", "Get a lift", "Motorcycle",
               "Taxi", "Bus", "Train", "Light rail",
               "Cycle", "Walk", "Other"))

df %>% 
  select(pidp, contains("wktrv"), -j_wktrvfar, -j_jswktrvfar) %>% 
  pivot_longer(cols = !pidp) %>% 
  mutate(name = str_remove(name, "j_js"),
         name = str_remove(name, "j_")) %>% 
  left_join(tran_opt) %>% 
  filter(value == 1) %>% 
  mutate(did_they = "Yes") %>% 
  select(-name, -value) %>% 
  pivot_wider(id_cols = pidp, names_from = val, values_from = did_they)
  



```


```{r data income}
df = df %>% 
  filter(! j_fimnnet_dv %in% c(-9, -8, -2, -1)) %>% 
  mutate(j_fimnnet_dv = as.numeric(j_fimnnet_dv))
```

```{r data dist to work}
df = df %>% 
  filter(j_workdis >= 0)
```





```{r data multi benefits, eval = F}
df %>% 
  select(pidp, j_benbase1, j_benbase2, j_benbase4) %>% 
  pivot_longer(!pidp) %>% 
  mutate(value = replace(value, value < 0, 0)) %>% 
  group_by(pidp) %>% 
  summarise(benefits = sum(value)) %>% 
  count(benefits)
```

## Age

```{r age, eval = F}
df %>% 
  count(j_pdvage) %>% 
  ggplot(aes(j_pdvage, n)) +
  geom_col() +
  labs(title = "Age of respondents",
       x = "Age",
       y = "Respondents") +
  theme_temp()
```



Employed:

```{r employed, eval = F}
df %>% 
  select(pidp, contains("j_wktrv"), -j_wktrvfar) %>% 
  pivot_longer(cols = !pidp) %>% 
  count(name, value) %>% 
  pivot_wider(names_from = name, values_from = n)
```

Self employed:

```{r self employed, eval = F}
df %>% 
  select(pidp, contains("j_jswktrv"), -j_jswktrvfar) %>% 
  pivot_longer(cols = !pidp) %>% 
  count(name, value) %>% 
  pivot_wider(names_from = name, values_from = n)
```

Larissa Pople notes we can add together travel from employed and self employed categories.

### How many people report multiple modes?

```{r multi-report, eval = F}
df %>% 
  select(pidp, contains("wktrv"), -j_wktrvfar, -j_jswktrvfar) %>% 
  pivot_longer(cols = !pidp) %>% 
  mutate(name = str_replace(name, "j_js", "j_")) %>% 
  filter(value == 1) %>% 
  count(pidp, name = "modes") %>% 
  count(modes, name = "respondents")
```


## Results

```{r distance mode}
x = df %>% 
  filter(j_wktrv1 == 1)

x %>% 
  ggplot(aes(j_workdis)) +
  stat_ecdf() +
  geom_vline(xintercept = 1.5, linetype = "dotted") +
  geom_vline(xintercept = 5, linetype = "dashed") +
  geom_vline(xintercept = 10, linetype = "dotdash") +
  coord_cartesian(xlim = c(0, 30)) +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "How many car commutes could be active travel?",
       subtitle = "Vertical lines of: walking, cycling, e-cycling.",
       x = "Distance to work",
       y = "Percent of respondents") +
  theme_temp()
```

* `r sum(x$j_workdis == 1)` of `r nrow(x)` (`r round(100 * sum(x$j_workdis == 1) / nrow(x))` %) of respondents drive to work is 1 mile :-(
* `r sum(x$j_workdis < 1.5)` of `r nrow(x)` (`r round(100 * sum(x$j_workdis < 1.5) / nrow(x))` %) of respondents drive to work is less than walking distance (1.5 miles)
* `r sum(x$j_workdis < 5)` of `r nrow(x)` (`r round(100 * sum(x$j_workdis < 5) / nrow(x))` %) of respondents drive to work is less than cycling distance (5 miles)
`r sum(x$j_workdis < 10)` of `r nrow(x)` (`r round(100 * sum(x$j_workdis < 10) / nrow(x))` %) of respondents drive to work is less than electric cycling distance (10 miles)


```{r income}
tran_opt = tibble(name = paste0("j_wktrv", c(1:10, 97)),
       val = c("Drive myself", "Get a lift", "Get a lift", "Motorcycle",
               "Taxi", "Bus", "Train", "Light rail",
               "Cycle", "Walk", "Other"))

x = df %>% 
  select(contains("j_wktrv"), j_fimnnet_dv) %>% 
  pivot_longer(!j_fimnnet_dv) %>% 
  filter(value == 1) %>% 
  left_join(tran_opt)

y = x %>% 
  group_by(val) %>% 
  summarise(median_in = median(j_fimnnet_dv))

x %>% 
  left_join(y) %>% 
  mutate(val = fct_reorder(val, median_in)) %>% 
  ggplot(aes(val, j_fimnnet_dv)) +
  geom_boxplot() +
  scale_y_log10() +
  labs(title = "What is the monthly income of different transport modes?",
       x = "",
       y = "Income (£, log scale)") +
  theme_temp()
```

