---
title: "Transport"
author: "Mike Spencer"
date: "25/11/2021"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = F, warning = F,
                      fig.width = 10, fig.height = 6)
options(scipen = 999)
```

```{r packages}
library(tidyverse)
library(haven)
library(knitr)
```

```{r plot theme}
theme_temp = function(){
  theme_bw() +
    theme(text = element_text(size = 15))
}
```


## Intro

This is an RMarkdown document, summarising variables of interest.
I'm considering the relationship between income and mode of travel.
For example, is cycling the preserve of the affluent middle class?

Understanding Society variable guide: <https://www.understandingsociety.ac.uk/documentation/mainstage/dataset-documentation?search_api_views_fulltext=salary>.

```{r data}
df = read_dta("~/Cloud/personal/gofcoe/understanding_society/6614stata_B17CC6790677EF32F72CE50881AE98E1B9FC1F79133B07B63B353396D3AB917A_V1/UKDA-6614-stata/stata/stata13_se/ukhls_w10/j_indresp.dta")
```

## Age

```{r age}
df %>% 
  count(j_pdvage) %>% 
  ggplot(aes(j_pdvage, n)) +
  geom_col() +
  labs(title = "Age of respondents",
       x = "Age",
       y = "Respondents") +
  theme_temp()
```


## Transport to work variables

### Main mode of transport to work

```{r commute main}
df %>% 
  count(j_wktrvfar)
```


### Mode(s) of transport for journey to work

Variables:

* wktrv1 	Drive myself by car or van
* wktrv2 	Get a lift with someone from household
* wktrv3 	Get a lift with someone outside the household
* wktrv4 	Motorcycle/moped/scooter
* wktrv5 	Taxi/minicab
* wktrv6 	Bus/coach
* wktrv7 	Train
* wktrv8 	Underground/Metro/Tram/Light railway
* wktrv9 	Cycle
* wktrv10 	Walk
* wktrv97 	Other


Response options:

* Value = 0.0	Label = Not mentioned
* Value = 1.0	Label = Yes mentioned
* Value = -1.0	Label = don't know
* Value = -9.0	Label = missing
* Value = -8.0	Label = inapplicable
* Value = -7.0	Label = proxy
* Value = -2.0	Label = refusal

Employed:

```{r employed}
df %>% 
  select(pidp, contains("j_wktrv"), -j_wktrvfar) %>% 
  pivot_longer(cols = !pidp) %>% 
  count(name, value) %>% 
  pivot_wider(names_from = name, values_from = n)
```

Self employed:

```{r self employed}
df %>% 
  select(pidp, contains("j_jswktrv"), -j_jswktrvfar) %>% 
  pivot_longer(cols = !pidp) %>% 
  count(name, value) %>% 
  pivot_wider(names_from = name, values_from = n)
```

Larissa Pople notes we can add together travel from employed and self employed categories.

### How many people report multiple modes?

```{r multi-report}
df %>% 
  select(pidp, contains("wktrv"), -j_wktrvfar, -j_jswktrvfar) %>% 
  pivot_longer(cols = !pidp) %>% 
  mutate(name = str_replace(name, "j_js", "j_")) %>% 
  filter(value == 1) %>% 
  count(pidp, name = "modes") %>% 
  count(modes, name = "respondents")
```


### Distance to work

```{r work dist}
df %>% 
  count(j_workdis)

df %>% 
  filter(j_workdis >= 0) %>% 
  ggplot(aes(j_workdis)) +
  geom_histogram() +
  scale_x_log10() +
  labs(title = "How far do respondents live from work?",
       x = "Distance (miles, log scale)",
       y = "Respondents")
```


## Individual income net

```{r individual income}
df %>% 
  filter(! j_fimnnet_dv %in% c(-9, -8, -2, -1)) %>% 
  transmute(j_fimnnet_dv = as.numeric(j_fimnnet_dv)) %>%
  ggplot(aes(j_fimnnet_dv)) +
  geom_histogram() +
  scale_x_log10() +
  labs(title = "Total net personal income",
       x = "Income (£, log scale)",
       y = "Respondents") +
  theme_temp()
```


## Benefits

### income support

```{r income support}
df %>% 
  count(j_benbase1)
```


### job seekers

```{r job seekers}
df %>% 
  count(j_benbase2)
```


### universal credit

```{r universal credit}
df %>% 
  count(j_benbase4)
```


### Multiple benefits

How many respondents claim one or multiple benefits, from income support, jobs seekers or universal credit.

```{r multi benefits}
df %>% 
  select(pidp, j_benbase1, j_benbase2, j_benbase4) %>% 
  pivot_longer(!pidp) %>% 
  mutate(value = replace(value, value < 0, 0)) %>% 
  group_by(pidp) %>% 
  summarise(benefits = sum(value)) %>% 
  count(benefits)
```

