---
title: "Transport mode analysis"
author: "Mike Spencer"
date: "09/12/2021"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = F, warning = F,
                      fig.width = 10, fig.height = 6)
options(scipen = 999)
```

```{r packages}
library(tidyverse)
library(haven)
library(knitr)
```

```{r plot theme}
theme_temp = function(){
  theme_bw() +
    theme(text = element_text(size = 15))
}
```


## Intro

This is an RMarkdown document, presenting and discussing analysis of transport mode data.
I'm considering the relationship between income and mode of travel.
For example, is cycling the preserve of the affluent middle class?

Understanding Society variable guide: <https://www.understandingsociety.ac.uk/documentation/mainstage/dataset-documentation?search_api_views_fulltext=salary>.


## Data prep

Only employed people considered, i.e. self employed excluded.
This is due to the increase in data complexity of including both types.

Non-commuting journeys are not available in the Understanding Society dataset.

```{r data}
df = read_dta("~/Cloud/personal/gofcoe/understanding_society/6614stata_B17CC6790677EF32F72CE50881AE98E1B9FC1F79133B07B63B353396D3AB917A_V1/UKDA-6614-stata/stata/stata13_se/ukhls_w10/j_indresp.dta") %>% 
  select(pidp, j_pdvage, j_sex,
         contains("j_wktrv"), j_workdis,
         j_fimnnet_dv,
         j_benbase1, j_benbase2, j_benbase4) %>% 
  mutate(j_sex = as_factor(j_sex))
```

```{r data transport}
tran_opt = tibble(name = paste0("j_wktrv", c(1:10, 97)),
       val = c("Drive myself", "Get a lift", "Get a lift", "Motorcycle",
               "Taxi", "Bus", "Train", "Light rail",
               "Cycle", "Walk", "Other"))
```

```{r data income}
df = df %>% 
  filter(! j_fimnnet_dv %in% c(-9, -8, -2, -1)) %>% 
  mutate(j_fimnnet_dv = as.numeric(j_fimnnet_dv))
```

```{r data dist to work}
df = df %>% 
  filter(j_workdis >= 0)
```

```{r data multi benefits}
x = df %>% 
  select(pidp, j_benbase1, j_benbase2, j_benbase4) %>% 
  pivot_longer(!pidp) %>% 
  mutate(value = replace(value, value < 0, 0)) %>% 
  group_by(pidp) %>% 
  summarise(benefits = sum(value))

df = df %>% 
  left_join(x)
```

```{r mode long}
df_long = df %>% 
  filter(j_pdvage > 0) %>% 
  select(contains("j_wktrv"), j_pdvage, j_sex, j_fimnnet_dv, j_workdis) %>% 
  pivot_longer(contains("j_wktrv")) %>% 
  filter(value == 1) %>% 
  left_join(tran_opt) %>% 
  filter(!is.na(val))
```


## Results

### Group sizes

```{r sizes table}
df_long %>% 
  count(val, j_sex) %>% 
  pivot_wider(names_from = j_sex, values_from = n) %>% 
  kable()
```

```{r sizes plot}
x = df_long %>% 
  count(val, name = "tot") %>% 
  mutate(lab = paste0(val, "\nn = ", tot))

y = df_long %>% 
  filter(j_sex == "female") %>% 
  count(val, name = "n_women")

df_long %>% 
  count(val, j_sex) %>% 
  left_join(x) %>% 
  left_join(y) %>% 
  mutate(pos = n_women / tot,
         lab = fct_reorder(lab, pos)) %>% 
  ggplot(aes(lab, n, fill = j_sex)) +
  geom_col(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "What is the proportion of each sex by transport type?",
       x = "",
       y = "Respondents",
       fill = "") +
  theme_temp() +
  theme(legend.position="bottom")
```


### Age and mode of transport

```{r age}
y = df_long %>% 
  group_by(val) %>% 
  summarise(median_in = median(j_pdvage),
            n = n()) %>% 
  mutate(lab = paste0(val, "\nn = ", n, "\nm = ", round(median_in)))

df_long %>% 
  left_join(y) %>% 
  mutate(lab = fct_reorder(lab, median_in)) %>% 
  ggplot(aes(lab, j_pdvage)) +
  geom_boxplot(size = 1.1) +
  geom_jitter(width = 0.2, alpha = 0.05) +
  labs(title = "What is the age of respondents using different transport modes?",
       subtitle = "n = number in group, m = median of group.",
       x = "",
       y = "Age (years)") +
  theme_temp()
```

* Younger people are more likely to get a lift to work. Getting a lift is an inherently dependent activity, which is asociated with being young.
* Motorbike commuting may increase in midlife ;-)


### Age, sex and mode of transport

```{r age and sex, fig.height=7}
y = df_long %>% 
  group_by(val) %>% 
  summarise(median_in = median(j_pdvage),
            n = n()) %>% 
  mutate(lab = paste0(val, "\nn = ", n, "\nm = ", round(median_in)))

df_long %>% 
  left_join(y) %>% 
  mutate(lab = fct_reorder(lab, median_in)) %>% 
  ggplot(aes(lab, j_pdvage, colour = j_sex)) +
  geom_boxplot(size = 1.1) +
  geom_jitter(width = 0.2, alpha = 0.05) +
  labs(title = "What is the age of respondents using different transport modes?",
       subtitle = "n = number in group, m = median of group.",
       x = "",
       y = "Age (years)",
       colour = "") +
  theme_temp() +
  theme(legend.position="bottom")
```


### Replaceable journeys

```{r distance mode, fig.height=7}
x = df %>% 
  filter(j_wktrv1 == 1)

x %>% 
  ggplot(aes(j_workdis, colour = j_sex)) +
  stat_ecdf(size = 1.2) +
  geom_vline(xintercept = 1.5, linetype = "dotted") +
  geom_vline(xintercept = 5, linetype = "dashed") +
  geom_vline(xintercept = 10, linetype = "dotdash") +
  coord_cartesian(xlim = c(0, 30)) +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "How many car commutes could be active travel?",
       subtitle = "Vertical lines of: walking, cycling, e-cycling.\nNote x axis is cropped.",
       x = "Distance to work (miles)",
       y = "Percent of respondents",
       colour = "") +
  theme_temp() +
  theme(legend.position="bottom")
```

* `r sum(x$j_workdis == 1)` of `r nrow(x)` (`r round(100 * sum(x$j_workdis == 1) / nrow(x))` %) of respondents drive to work is 1 mile :-(
* `r sum(x$j_workdis < 1.5)` of `r nrow(x)` (`r round(100 * sum(x$j_workdis < 1.5) / nrow(x))` %) of respondents drive to work is less than walking distance (1.5 miles)
* `r sum(x$j_workdis < 5)` of `r nrow(x)` (`r round(100 * sum(x$j_workdis < 5) / nrow(x))` %) of respondents drive to work is less than cycling distance (5 miles)
* `r sum(x$j_workdis < 10)` of `r nrow(x)` (`r round(100 * sum(x$j_workdis < 10) / nrow(x))` %) of respondents drive to work is less than electric cycling distance (10 miles)
* More men travel further to work
* What are the barriers to women undertaking short journeys sustainably? Is this linked to trip chaining, unsafe routes, something else?


### Income and mode of transport

```{r income}
y = df_long %>% 
  group_by(val) %>% 
  summarise(median_in = median(j_fimnnet_dv),
            n = n()) %>% 
  mutate(lab = paste0(val, "\nn = ", n, "\nm = ", round(median_in)))

df_long %>% 
  left_join(y) %>% 
  mutate(lab = fct_reorder(lab, median_in)) %>% 
  ggplot(aes(lab, j_fimnnet_dv)) +
  geom_boxplot(outlier.alpha = 0, size = 1.1) +
  geom_jitter(width = 0.2, alpha = 0.05) +
  coord_cartesian(ylim = c(0, 4000)) +
  labs(title = "What is the monthly income of different transport modes?",
       subtitle = "n = number in group, m = median of group.\nNote y axis is cropped.",
       x = "",
       y = "Income (£)") +
  theme_temp()
```

* 3 step changes:
  * Get a lift is a dependent activity
  * Driving/cycling/motor cycling more automous than bus/taxi travel
  * Light rail and train access to more prestigious, city centre jobs.
* This type of analysis may support transport poverty work - areas with less transport links reduce job options.

```{r income sex, fig.height=7}
y = df_long %>% 
  group_by(val) %>% 
  summarise(median_in = median(j_fimnnet_dv),
            n = n()) %>% 
  mutate(lab = paste0(val, "\nn = ", n, "\nm = ", round(median_in)))

df_long %>% 
  left_join(y) %>% 
  mutate(lab = fct_reorder(lab, median_in)) %>% 
  ggplot(aes(lab, j_fimnnet_dv, colour = j_sex)) +
  geom_boxplot(outlier.alpha = 0, size = 1.1) +
  geom_jitter(width = 0.2, alpha = 0.05) +
  coord_cartesian(ylim = c(0, 4000)) +
  labs(title = "What is the monthly income of different transport modes?",
       subtitle = "n = number in group, m = median of group.\nNote y axis is cropped.",
       x = "",
       y = "Income (£)",
       colour = "") +
  theme_temp() +
  theme(legend.position="bottom")
```


### Relationships

```{r income and age, fig.height = 8}
df_long %>% 
  filter(value == 1) %>% 
  ggplot(aes(j_pdvage, j_fimnnet_dv)) +
  geom_point(alpha = 0.15) +
  stat_smooth() +
  facet_wrap(~val) +
  scale_y_continuous(labels = scales::comma) +
  labs(title = "How does income relate to age?",
       x = "Age (years)",
       y = "Monthly income (£)") +
  theme_temp()
```

```{r income and distance, fig.height = 8}
df_long %>% 
  filter(value == 1) %>% 
  ggplot(aes(j_workdis, j_fimnnet_dv)) +
  geom_point(alpha = 0.15) +
  stat_smooth() +
  facet_wrap(~val) +
  scale_y_continuous(labels = scales::comma) +
  coord_cartesian(xlim = c(0, 100)) +
  labs(title = "How does distance to work relate to age?",
       subtitle = "Distance axis cropped. Some people may use multiple transport modes",
       x = "Distance (miles)",
       y = "Monthly income (£)") +
  theme_temp()
```

* On average, people seem to travel further for better paid work.
* In practice, I think people are unwilling to travel further for less well paid work. i.e. income does seem to increase with greater distance, but lower incomes drop out.
